---
title: "Problem Set: Vectors, Strings, Factors, and Dates"
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(nycflights13)
library(lubridate)
```

------------------------------------------------------------------------

## Question 1 — Exploring the Star Wars Universe

The `dplyr::starwars` dataset (available as soon as you load `tidyverse`) contains physical and biographical information on characters from the Star Wars films. Start by running `View(starwars)` and ?starwars to get oriented.

The code below stores the characters with blue eyes in the "blue_eyes" dataframe, and the characters with red eyes in the "red_eyes" dataframe. There is one character with both blue and red eyes. Use a logical vector approach again with str_detect to find out which one. Hint: one way to do it is with the logical & operator, which returns all rows for which multiple criteria are met

```{r}
# Characters with "blue" in eye_color
blue_eyes <- starwars |>
  filter(str_detect(eye_color, "blue"))
blue_eyes

# Characters with "brown" in eye_color
red_eyes <- starwars |>
  filter(str_detect(eye_color, "red"))
red_eyes

blue_and_red <- red_eyes |> 
  filter(str_detect(eye_color, "blue"))
#R4-P17
```

## Question 2 — American Unemployment Over Time

The `ggplot2::economics` dataset contains monthly U.S. macroeconomic indicators from 1967 to 2015. The `date` column is already a proper `Date` object.

**(a)** The goal here is to see how the unemployment rate changes over time. Start by plotting the number of unemployed people over time:

```{r}
  ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line()
```

Now use lubridate to plot only the unemployment data from Jan 1st 2008 to Jan 1st 2009. Hint: use the lubridate function ymd to specify date windows, and use a similar logical criteria to in Q1, above

```{r}
# your code here

economics <- ggplot2::economics
smaller_range <- ymd("2008-01-01")%--%(ymd("2009-01-01"))

df <- economics |>
  filter(date>ymd("2007-12-01")) |>
  filter(date<ymd("2009-02-01"))


ggplot(df, aes(x = date, y = unemploy)) +
  geom_line()
```

What do you notice about the format of the date variable on the x axis of this plot vs the plot over the entire time window. What do you think is happening here?

**Your answer:** In the first plot, only the year is shown. In the second plot, both month and year are shown. I think this is because R was told the date format in the second plot, so it knows how to correctly format the month.

Now use mutate to calculate a new variable - the monthly unemployment rate as `unemploy / pop * 100`. Make a plot over the whole time period as in part 1 of this question.

```{r}
# your code here

new_economics <- economics |>
  mutate(unemp_rate=(unemploy / pop * 100))

ggplot(new_economics, aes(x = date, y = unemp_rate)) +
  geom_line()
```

Do you draw different conclusions now about how unemployment has changed over these 40 years? what explains the difference?

**your answer:** Yes. Both graphs show different increases and decreases in their respective y-axis. This is because the first graph is showing only the number of unemployed, while the last shows the rate, which takes into account the overall changes in population.

## Question 3 — Flight Status and the Value of Factors

Run the code below to create a `status` variable that categorises each flight by how early or late it arrived.

```{r}
flight_status <- flights |>
  mutate(
    status = case_when(
      is.na(arr_delay)      ~ "cancelled",
      arr_delay < -30       ~ "very early",
      arr_delay < -15       ~ "early",
      abs(arr_delay) <= 15  ~ "on time",
      arr_delay < 60        ~ "late",
      arr_delay < Inf       ~ "very late",
    ),
    .keep = "used"
  )
```

**(a)** Without converting `status` to a factor, run the followong code to create a bar chart showing the count of flights in each category. What do you notice about the ordering of the bars? Is it meaningful?

```{r}
ggplot(flight_status, aes(x = status)) +
  geom_bar()
```

**your answer:** The order of the bars don't follow a logical order from earliest to latest. It is in alphabetical order instead.

**(b)** Convert `status` to a factor with levels in a natural severity order: `"very early"`, `"early"`, `"on time"`, `"late"`, `"very late"`, `"cancelled"`. Save it back into `flight_status` and recreate the bar chart. How does the chart change?

```{r}
# your code here

status_order <- c("very early","early","on time","late","very late", "cancelled")
new_status <- factor(status, levels=status_order)

ggplot(flight_status, aes(x = factor(status, levels=status_order))) +
  geom_bar()
```

**your answer:** The chart is now ordered by severity of earliest to latest.
